### 2020-09-08 ART_Mscislawski/Mscislaswki-PU.tex
Authors' and editorial corrections added

\sectionno{Zakończenie}
\noindent Nie sposób oprzeć się wrażeniu, że Bohdan Szyszkowski jest myślicielem bardzo oryginalnym, ale też -- zwłaszcza jeśli chodzi o~kijowski okres jego refleksji filozoficznej -- poniekąd nieznanym. Zaskakująca może być próba połączenia w~jednym, filozoficznym opisie problematyki związanej z~materią, duchem i~życiem, jednakże bez próby zastosowania prymitywnego redukcjonizmu, jak sugeruje treść jego wykładów ,,Z zagadnień filozofji przyrody''. Artykuł \textit{O~ciągłości} ukazuje Szyszkowskiego jako myśliciela wyczulonego na subtelne niuanse, czego świadectwem jest wspomniane już odróżnienie struktur matematycznych od opisywanych przez nie obiektów fizycznych. Interesującą propozycją, wspartą interesującą argumentacją, jest także jego postulat istnienia eteru jako gwarancji utrzymania ciągłości jako pojęcia fundamentalnego -- co w~jego opinii -- pozwala zachować kategorię przyczynowości. Trzeba jednak w~tym miejscu podkreślić zasadniczą trudność w~odniesieniu do tezy polskiego chemika: czy intuicyjne pojęcie ciągłości, które stosuje on do umysłu, jest tym samym pojęciem, którym operuje się w~matematyce? Jak należałoby rozumieć ciągłość eteru, który nie jest ani bytem mentalnym, ani -- tym bardziej -- strukturą matematyczną? Są to pytania, na które Szyszkowski nie udziela odpowiedzi. Można także odnieść wrażenie, że ma on trafne intuicje, np. pojęcia eteru, stwierdzając, że potrzebne medium o~własnościach fizycznych, w~którym może mieć miejsce propagacja fenomenów fizycznych (pojęcie analogiczne do współczesnego pojęcia pola). Nie potrafi odpowiednio tych intuicji wyrazić, być może właśnie poprzez brak odpowiedniej bazy pojęciowej. Niekiedy też intuicje te zdają się ginąć w~całokształcie myśli polskiego chemika, mającego ambicję stworzyć ogólny ogląd rzeczywistości. W~sposób szczególny dotyczy to treści wspomnianych wyżej wykładów. Charakterystyczną cechą spostrzeżeń Szyszkowskiego, utrudniającą ich analizę, jest także to, że często fundamentalne dla jego analiz stwierdzenia przytacza on bez jakichkolwiek argumentów czy choćby śladowych uzasadnień. Dotyczy to -- przykładowo -- tak istotnego stwierdzenia, iż ciągłość jest wrodzonym pojęciem umysłu. Szyszkowski nie podaje -- niestety -- jakiegokolwiek uzasadnienia takiej tezy, która wydaje się być istotna dla jego poglądów
%\label{ref:RND9xEAwphLhJ}(por. Szyszkowski, 1916, s.~47).
\parencite[por.][s.~47]{szyszkowski_o_1916}. %
Pozostaje pytaniem otwartym, na ile poglądy Szyszkowskiego są oryginalne, a~na ile inspirowane przez francuski konwencjonalizm, zwłaszcza zaś przez idee Henri
Poincarégo\footnote{Na temat powiązania równań różniczkowych (oraz ciągłością w~sensie matematycznym) z~determinizmem por. 
%\label{ref:RNDuHKnV5BcOv}(Poincaré, 1908a, s.~104–132, b, s.~98–135, 160–169).
\parencites[][s.~98–135]{poincare_nauka_1908}[][s.~98–135, 160–169]{poincare_wartosc_1908}. O~tej kwestii zob. także 
%\label{ref:RNDo02BSNCiJ4}(Leszczyński, Szlachcic, 2003, s.~80–81).
\parencite[][s.~80–81]{leszczynski_wprowadzenie_2003}.}.

>>>>

\sectionno{Zakończenie}
\noindent Nie sposób oprzeć się wrażeniu, że Bohdan Szyszkowski jest myślicielem bardzo oryginalnym, ale też -- zwłaszcza jeśli chodzi o~kijowski okres jego refleksji filozoficznej -- poniekąd nieznanym. Zaskakująca może być próba połączenia w~jednym, filozoficznym opisie problematyki związanej z~materią, duchem i~życiem, jednakże bez próby zastosowania prymitywnego redukcjonizmu, jak sugeruje treść jego wykładów ,,Z zagadnień filozofji przyrody''. Artykuł \textit{O~ciągłości} ukazuje Szyszkowskiego jako myśliciela wyczulonego na subtelne niuanse, czego świadectwem jest wspomniane już odróżnienie struktur matematycznych od opisywanych przez nie obiektów fizycznych. Interesującą propozycją, wspartą ciekawą argumentacją, jest także jego postulat istnienia eteru jako gwarancji utrzymania ciągłości jako pojęcia fundamentalnego -- co w~jego opinii -- pozwala zachować kategorię przyczynowości. Trzeba jednak w~tym miejscu podkreślić zasadniczą trudność w~odniesieniu do tezy polskiego chemika: czy intuicyjne pojęcie ciągłości, które stosuje on do umysłu, jest tym samym pojęciem, którym operuje się w~matematyce? Jak należałoby rozumieć ciągłość eteru, który nie jest ani bytem mentalnym, ani -- tym bardziej -- strukturą matematyczną? Są to pytania, na które Szyszkowski nie udziela odpowiedzi. Można także odnieść wrażenie, że ma on trafne intuicje, np. pojęcia eteru, stwierdzając, że potrzebne medium o~własnościach fizycznych, w~którym może mieć miejsce propagacja fenomenów fizycznych (pojęcie analogiczne do współczesnego pojęcia pola). Nie potrafi odpowiednio tych intuicji wyrazić, być może właśnie poprzez brak odpowiedniej bazy pojęciowej. Niekiedy też intuicje te zdają się ginąć w~całokształcie myśli polskiego chemika, mającego ambicję stworzyć ogólny ogląd rzeczywistości. W~sposób szczególny dotyczy to treści wspomnianych wyżej wykładów. Charakterystyczną cechą spostrzeżeń Szyszkowskiego, utrudniającą ich analizę, jest także to, że często fundamentalne dla jego analiz stwierdzenia przytacza on bez jakichkolwiek argumentów czy choćby śladowych uzasadnień. Dotyczy to -- przykładowo -- tak istotnego stwierdzenia, iż ciągłość jest wrodzonym pojęciem umysłu. Szyszkowski nie podaje -- niestety -- jakiegokolwiek uzasadnienia takiej tezy, która wydaje się być istotna dla jego poglądów
%\label{ref:RND9xEAwphLhJ}(por. Szyszkowski, 1916, s.~47).
\parencite[por.][s.~47]{szyszkowski_o_1916}. %
Pozostaje pytaniem otwartym, na ile poglądy Szyszkowskiego są oryginalne, a~na ile inspirowane przez francuski konwencjonalizm, zwłaszcza zaś przez idee Henri
Poincarégo\footnote{Na temat powiązania równań różniczkowych (oraz ciągłością w~sensie matematycznym) z~determinizmem por. 
%\label{ref:RNDuHKnV5BcOv}(Poincaré, 1908a, s.~104–132, b, s.~98–135, 160–169).
\parencites[][s.~98–135]{poincare_nauka_1908}[][s.~98–135, 160–169]{poincare_wartosc_1908}. O~tej kwestii zob. także 
%\label{ref:RNDo02BSNCiJ4}(Leszczyński, Szlachcic, 2003, s.~80–81).
\parencite[][s.~80–81]{leszczynski_wprowadzenie_2003}.}.

### 2020-09-08 ART_Szybka/Szybka-PU.tex
Authors' and editorial corrections added

\begin{artengenv}{Sebastian Szybka}
	{Some remarks on the first image of a black hole}
	{Some remarks on the first image of a black hole}
	{Some remarks on the first image of a black hole}
	{Astronomical Observatory, Jagiellonian University\\
	Copernicus Center for Interdisciplinary Studies}
	{On the 10\textsuperscript{th} of April, 2019 the Event Horizon Telescope Collaboration presented the first image of the black hole. The image was obtained with a planet-scale array of eight ground-based radio telescopes. The observation relied on a technique called very long base interferometry which synchronises telescope facilities around the world. The image of a black hole together with the recent detections of gravitational waves confirms one of the most intriguing predictions of Einstein's gravity theory, namely, the existence of black holes. I will provide more details on this remarkable observation and explore its consequences for our understanding of nature. The physical reality of black holes is strongly supported by recent advances of astronomy. I claim that this fact is the key to understanding the relation between our world and the world of mathematics.}
	{black holes, event horizon, general relativity, interferometry, astronomy, Einstein, geometry.}

@@

How to create an image of a black hole? Since the event horizon of the M87 black hole is about 22 micro-arc-sec of a degree as seen from Earth, an extreme resolution is needed. A resolution of a human eye is about 60 arc-sec of a degree. The Hubble telescope is much better having a resolution about 0.05 arc-sec of a degree, but still nowhere close to what is needed. We need an instrument which would allow a person sitting in Kraków to read a newspaper in New York! A simple rule says: the larger diameter of the lens the higher resolution. The formula which is necessary to estimate the size of the instrument has a trivial form. The resolution of an astronomical instrument is about $\lambda/D$ where $\lambda$ is the wavelength of the electromagnetic radiation at which we want to observe and $D$ is the size of the instrument. Firstly, the Event Horizon Telescope team had to decide at which wavelength they want to observe a black hole. Several factors must be taken into account here: longer wavelengths require bigger instruments, the M87 black hole should look interesting at a given wavelength, wavelengths at which the Earth-based astronomical observations are conducted are strictly restricted by absorption of the Earth's atmosphere. Majority of people would like to have a photograph of a black hole and by that they usually mean: in the visible light. However, the wavelength of the electromagnetic waves is not their fundamental property---it depends on the observer who measures them. There is no fundamental difference between photons associated to the visible light and those associated to radio waves. The techniques used to capture them differs because these photons have different energies relative to us, but what is infrared radiation for one observer may be visible light to another observer who moves relative to the first one at relativistic speed. Therefore, for a professional astronomer the image of the black hole in different part of the electromagnetic spectrum than the visible range is not less `true' nor interesting than a photograph in the visible light. (Even cameras in ordinary mobile phones go slightly beyond visible part of the spectrum.) Moreover, it was expected that our black holes do not look attractive from this distance in the visible light---they would probably look more like an ordinary stars and we would not see a black hole shadow. Observations in standard radio waves were not convenient because necessary resolution would imply too large size of the instruments (from formula $\lambda/D$). It turned out that the wavelength $1.3$ mm is the best choice. It was expected that we will see the black hole shadow, there is the so-called `atmospheric window' at this wavelength (a tiny one), and, what a lucky coincidence, the M87 black hole is just at the right distance, hence, the instrument has largest admissible size---the size of Earth!

@@

Imagine that the dish of our satellite antenna has a hole. It will still work, but less efficiently. The size of the astronomical instrument $D$ that enters the formula for a resolution $\lambda/D$ denotes the distance between most distant parts of our instrument. Therefore, a dish with many holes would have the same resolution as the dish without holes, but the signal would be much weaker. The basic idea of the VLBI technique is to use many different instruments spread across the globe and integrate the signal. Thus, all of them may work as a single instrument. The clear advantage of this method is that there is no need to build a new instrument and one may use pre-existing infrastructure. The basics of this technique has been established many years ago, but only recently the technical difficulties were overcome. The highly non-trivial task is to correlate the signal between different instruments. One should know with extremely high precision when the signal arrived. Clocks at different observatories must be precise and synchronized. A delay of the signal that is usually induced by the telescope electronics, must be estimated to high accuracy. All these goals were achieved by the Event Horizon Telescope team. The atomic clocks used to time-stamp data were hydrogen masers. These clocks lose only one second in every 100 million years. The required synchronisation obtained with Global Positioning System was at the level of a millionth of a second.

The signal from the black hole which the Event Horizon Telescope Collaboration was going to observe was weak. In fact, it was billion times weaker than the electromagnetic signal received by old fashion TV antenna. Nevertheless, the array of eight radio telescopes (ALMA, APEX, the IRAM 30-meter telescope, the James Clerk Maxwell Telescope, the Large Millimeter Telescope Alfonso Serrano, the Submillimeter Array, the Submillimeter Telescope, and the South Pole Telescope) was powerful enough to observe the M87 black hole during a week long session. (The Event Horizon Telescope project has been in the works for two decades to make this observations possible.) The radio telescopes do not create images as `single shots.' They rather work like scanners which measure intensity of the radiation at a given point. The instruments recorded hundreds of petabytes of data which was combined into a single image by supercomputers at the Max Planck Institute for Radio Astronomy and MIT Haystack Observatory. (What is interesting, it was faster to transport hard disks with data on planes than send data in a digital form over internet.)

@@

The image of the black hole at the center of the galaxy M87 is presented in figure \ref{bh}. The black hole was observed at a single wavelength $1.3$ mm, thus the gradation in color  does not corresponds to different wavelengths, but to the brightness (colors are artificial). The matter falling onto black hole forms an accretion disk. It is heated and glows. The gravity bends light which forms a bright ring around a dark shadow of the black hole. For a non-spinning black hole (but this black hole is spinning) the inner radius of the ring is about $2.6$ times bigger then the radius of the event horizon. At first sight, it looks like we see the M87 black hole in the direction perpendicular to the disk. This is not true. The light rays are bend so strongly that we see also the part of the accretion disk which should be hidden behind the black hole. One part of the ring is brighter, because of the so-called relativistic beaming effect: the increased brightness implies that this part of the accretion disk rotates towards us.
%
\begin{figure}[t!]
\begin{center}
\includegraphics[width=\textwidth]{ART_Szybka/szybka-imgbw.png}
\caption{The image of the supermassive black hole at the center of the galaxy M87 (the black hole is $6.5$ billion times more massive than the Sun). The matter falling on the black hole shines radiating energy away. The extreme gravity curves light which forms a bright ring in the image. At the center of the ring a dark silhouette of the black hole is clearly visible. This shadow contains the event horizon. Credit: Event Horizon Telescope Collaboration.}
\label{bh}
\end{center}
\end{figure}
%

@@

The first image of the black hole (figure \ref{bh}) is consistent with the general relativity. Of course, there are many other images that would be consistent with the Einstein's gravity theory (depending on the parameters of the system), thus in this case it is better to talk about consistency than a direct proof of the theory. Nevertheless, many characteristic features of this image agree with predictions of general relativity and one may use it to improve estimates of the M87 black hole parameters (like a mass and a size). With a help of this image we may exclude some exotic alternatives to black holes.

>>>>

\begin{artengenv}{Sebastian J. Szybka}
	{Some remarks on the first image of a black hole}
	{Some remarks on the first image of a black hole}
	{Some remarks on the first image of a black hole}
	{Astronomical Observatory, Jagiellonian University\\
	Copernicus Center for Interdisciplinary Studies}
	{On the 10\textsuperscript{th} of April, 2019 the Event Horizon Telescope Collaboration presented the first image of the black hole. The image was obtained with a planet-scale array of eight ground-based radio telescopes. The observation relied on a technique called very long base interferometry which synchronises telescope facilities around the world. The image of a black hole together with the recent detections of gravitational waves confirms one of the most intriguing predictions of Einstein's gravity theory, namely, the existence of black holes. I will provide more details on this remarkable observation and explore its consequences for our understanding of nature. The physical reality of black holes is strongly supported by recent advances of astronomy. I claim that this fact is the key to understanding the relation between our world and the world of mathematics.}
	{black holes, event horizon, general relativity, interferometry, astronomy, Einstein, geometry.}

@@

How to create an image of a black hole? Since the event horizon of the M87 black hole is about 22 micro-arc-sec of a degree as seen from Earth, an extreme resolution is needed. A resolution of a human eye is about 60 arc-sec of a degree. The Hubble telescope is much better having a resolution about 0.05 arc-sec of a degree, but still nowhere close to what is needed. We need an instrument which would allow a person sitting in Kraków to read a newspaper in New York! A simple rule says: the larger diameter of the lens the higher resolution. The formula which is necessary to estimate the size of the instrument has a trivial form. The resolution of an astronomical instrument is about $\lambda/D$ where $\lambda$ is the wavelength of the electromagnetic radiation at which we want to observe and $D$ is the size of the instrument. Firstly, the Event Horizon Telescope team had to decide at which wavelength they want to observe a black hole. Several factors must be taken into account here: longer wavelengths require bigger instruments, the M87 black hole should look interesting at a given wavelength, wavelengths at which the Earth-based astronomical observations are conducted are strictly restricted by absorption of the Earth's atmosphere. Majority of people would like to have a photograph of a black hole and by that they usually mean: in the visible light. However, the wavelength of the electromagnetic waves is not their fundamental property---it depends on the observer who measures them. There is no fundamental difference between photons associated to the visible light and those associated to radio waves. The techniques used to capture them differs because these photons have different energies relative to us, but what is infrared radiation for one observer may be visible light to another observer who moves relative to the first one at relativistic speed. Therefore, for a professional astronomer the image of the black hole in different part of the electromagnetic spectrum than the visible range is not less `true' nor interesting than a photograph in the visible light. (Even cameras in ordinary mobile phones go slightly beyond visible part of the spectrum.) Moreover, it was expected that our black holes do not look attractive from this distance in the visible light---they would probably look more like an ordinary stars and we would not see a black hole shadow. Observations in standard radio waves were not convenient because necessary resolution would imply too large size of the instruments (from formula $\lambda/D$). It turned out that the wavelength $1.3$ mm is the best choice: it was expected that we will see the black hole shadow, there is the so-called `atmospheric window' at this wavelength (a tiny one), and, what a lucky coincidence, the M87 black hole is just at the right distance, hence, the instrument has the largest admissible size---the size of Earth!

@@

Imagine that the dish of our satellite antenna has a hole. It will still work, but less efficiently. The size of the astronomical instrument $D$ that enters the formula for a resolution $\lambda/D$ denotes the distance between most distant parts of our instrument. Therefore, a dish with many holes would have the same resolution as the dish without holes, but the signal would be much weaker. The basic idea of the VLBI technique is to use many different instruments spread across the globe and integrate the signal. Thus, all of them may work as a single instrument. The clear advantage of this method is that there is no need to build a new instrument and one may use pre-existing infrastructure. The basics of this technique has been established many years ago, but only recently the technical difficulties were overcome. The highly non-trivial task is to correlate the signal between different instruments. One should know with extremely high precision when the signal arrived. Clocks at different observatories must be precise and synchronized. A delay of the signal that is usually induced by the telescope electronics, must be estimated to high accuracy. All these goals were achieved by the Event Horizon Telescope team. The atomic clocks used to time-stamp data were hydrogen masers. These clocks lose only one second in every 100 million years. The required synchronisation obtained with the Global Positioning System was at the level of a millionth of a second.

The signal from the black hole which the Event Horizon Telescope Collaboration was going to observe was weak. In fact, it was billion times weaker than the electromagnetic signal received by an old-fashion TV antenna. Nevertheless, the array of eight radio telescopes (ALMA, APEX, the IRAM 30-meter telescope, the James Clerk Maxwell Telescope, the Large Millimeter Telescope Alfonso Serrano, the Submillimeter Array, the Submillimeter Telescope, and the South Pole Telescope) was powerful enough to observe the M87 black hole during a week long session. (The Event Horizon Telescope project has been in the works for two decades to make this observations possible.) The radio telescopes do not create images as `single shots.' They rather work like scanners which measure intensity of the radiation at a given point. The instruments recorded hundreds of petabytes of data which was combined into a single image by supercomputers at the Max Planck Institute for Radio Astronomy and MIT Haystack Observatory. (What is interesting, it was faster to transport hard disks with data on planes than send data in a digital form over internet.)

@@

The image of the black hole at the center of the galaxy M87 is presented in figure \ref{bh}. The black hole was observed at a single wavelength $1.3$ mm, thus the gradation in color\edtfootnote{The printed format of this article contains black-and-white graphics, the electronic format is prepared in color.} does not corresponds to different wavelengths, but to the brightness (colors are artificial). The matter falling onto black hole forms an accretion disk. It is heated and glows. The gravity bends light which forms a bright ring around a dark shadow of the black hole. For a non-spinning black hole (but this black hole is spinning) the inner radius of the ring is about $2.6$ times bigger then the radius of the event horizon. At first sight, it looks like we see the M87 black hole in the direction perpendicular to the disk. This is not true. The light rays are bend so strongly that we see also the part of the accretion disk which should be hidden behind the black hole. One part of the ring is brighter, because of the so-called relativistic beaming effect: the increased brightness implies that this part of the accretion disk rotates towards us.
%
\begin{figure}[t!]
\begin{center}
\includegraphics[width=\textwidth]{ART_Szybka/szybka-imgbw.png}
\caption{The image of the supermassive black hole at the center of the galaxy M87 (the black hole is $6.5$ billion times more massive than the Sun). The matter falling on the black hole shines radiating energy away. The extreme gravity curves light which forms a bright ring in the image. At the center of the ring a dark silhouette of the black hole is clearly visible. This shadow contains the event horizon. Credit: Event Horizon Telescope Collaboration.}
\label{bh}
\end{center}
\end{figure}
%

@@

The first image of the black hole (figure \ref{bh}) is consistent with general relativity. Of course, there are many other images that would be consistent with the Einstein's gravity theory (depending on the parameters of the system), thus in this case it is better to talk about consistency than a direct proof of the theory. Nevertheless, many characteristic features of this image agree with predictions of general relativity and one may use it to improve estimates of the M87 black hole parameters (like a mass and a size). With a help of this image we may exclude some exotic alternatives to black holes.

### 2020-09-08 ART_Tambor_Szydlowski/Tambor_Szydlowski-PU.tex
Authors' and editorial corrections added

\begin{artplenv2auth}{Paweł Tambor, Marek Szydłowski}
	{Czy fizyczne teorie efektywne są adekwatną strategią osiągnięcia teorii ostatecznej?}
	{Czy fizyczne teorie efektywne są adekwatną strategią\ldots}
	{Czy fizyczne teorie efektywne są adekwatną strategią osiągnięcia teorii ostatecznej?}
	{Wydział Filozofii KUL, Uniwersytet Jagielloński Kraków}
	{Are Physical Effective Theories the Reliable Strategy for Achieving Certain Knowledge?}
	{The paper presents the methodology of effective theories as a~strategy used in the process of development of modern physics to reach a~final theory. We present the definition and characteristic features of an effective theory, as well as the answer to the question of whether and what kind of scenario of reaching a~final theory is realized by contemporary physics. We argue that the process of development of physics in the direction of a~final theory is potentially final, i.e. expressible in the conceptual schema of effective theories and as such it is convergent to a~final theory. In each effective theory there are physical constants, however, whose status differs from logical constants. They have a~dimension (length, energy, etc.) and are used to compare physical quantities. The structure of relevant effective theory can be interpreted in the epistemological framework of approximated truth theory. In the case study of cosmological models the sequence of models is convergent to potentially true model. The Standard Cosmological Model is the theory of the structure and dynamics of the Universe.}
	{philosophy and methodology of science, effective theory, final theory, cosmology, approximated truth theories.}
	{%
			{\flushright\subbold{Paweł Tambor}\\\subsubsectit\small{Katolicki Uniwersytet Lubelski}\par}%
			{\flushright\subbold{Marek Szydłowski}\\\subsubsectit\small{Uniwersytet Jagielloński}\par}%
		}

@@

\section{Nomenklatura: teorie efektywne, fundamentalne, ostateczne i~relacje między nimi}
Teoria efektywna posiada przede wszystkim następujące cechy\footnote{Prototypem teorii efektywnej na gruncie fizyki jest efektywna teoria pola
%\label{ref:RNDdsViEs8BX9}(Kim, 1998; Wilson, 1971).
\parencites[][]{kim_wilson_1998}[][]{wilson_renormalization_1971}. %
 W~niniejszej pracy podejmujemy próbę adaptacji pojęcia ,,teoria efektywna'' w~kontekście metodologii i~filozofii nauki.}:
 a)~wyraźnie określony i~ograniczony (najczęściej w~skali energetycznej) zakres stosowania; b) posługuje się parametrami, których natury sama teoria efektywna nie wyjaśnia, co nie przeszkadza w~skutecznym posługiwaniu się teorią; c) posługuje się pojęciami, które mają w~jej ramach status użytecznych fikcji (ze względu na zastosowane idealizacje); d) nie da się do niej stosować predykatu prawdziwości, ze względu na zastosowane idealizacje\footnote{Cechy (c) i~(d) są specyficzne dla analizy Uzana i~nie są powszechnie przyjmowane w~literaturze}. Jean-Philippe Uzan potwierdza taką intuicję dotyczącą teorii efektywnych; a~zwłaszcza to, że standardowy model kosmologiczny jest taką~właśnie teorią. Model ten z~punktu widzenia teorii fizyki opiera się~na ogólnej teorii względności, elektromagnetyzmie i~fizyce jądrowej. Dość~dobrze da się~wyodrębnić jego ograniczenia w~skali energii. A~zatem, stwierdza Uzan, żeby dobrze rozumieć w~przyjętej skali nasz obserwowalny Wszechświat, nie musimy konstruować kwantowej teorii grawitacji 
%\label{ref:RNDtodLlEAyWN}(Uzan, 2017, s.~109).
\parencite[][s.~109]{uzan_emergent_2017}.%

@@

Obok zaprezentowania kategorii teorii efektywnych i~ostatecznych, celem naszej pracy jest pewna spekulacja nad możliwymi scenariuszami poszukiwania teorii ostatecznej i~propozycja oparcia tych strategii na pojęciu ciągu teorii efektywnych. W~przyjętym rozumieniu teorii ostatecznej, będzie to proces osiągnięcia teorii ostatecznej w~sensie cząstkowym. Cytowany Szynkiewicz w~kontekście własnych analiz wskazuje za Barrowem na cztery scenariusze rozwoju nauki, traktowanej w~aspekcie treściowym
%\label{ref:RNDwI0uwM5FZZ}(Szynkiewicz, 2009, s.~118).
\parencite[][s.~118]{szynkiewicz_teorie_2009}. %
 W~pierwszym (oznaczonym predykatami: natura nieograniczona/ zdolności poznawcze nieograniczone) nauka rozwija się w~sposób nieograniczony. W~tym scenariuszu nie ma mowy o~osiągnięciu wiedzy, którą nazwalibyśmy teorią/teoriami ostatecznymi. W~drugim (natura nieograniczona/ zdolności poznawcze ograniczone) mamy do czynienia z~dwoma kierunkami rozwoju nauki: a) z~ograniczonymi zdolnościami poznawczymi związany może być (choć nie implikowany przez nie) także nieskończony rozwój nauki, a~zatem nie ma teorii ostatecznej, b) teoria ostateczna jest nieosiągalna, ale z~tego powodu, że ograniczone zdolności poznawcze (także ze względów ekonomicznych lub technicznej niezdolności przetworzenia ogromnej liczby danych) spowodują, że ostatecznie \textit{skazani} będziemy na teorie efektywne. Trzeci scenariusz rozwoju nauki (natura ograniczona/ zdolności poznawcze nieograniczone), jest optymistyczny i~przewiduje możliwość dopracowania się teorii ostatecznej lub przynajmniej zbioru teorii fundamentalnych. Pogląd ten wspierają sukcesy na polu konstruowania teorii kwantowych. Czwarta wizja rozwoju nauki wychodzi z~założenia o~ograniczoności samej natury i~ograniczoności zdolności poznawczych. Możliwe scenariusze zależą od tego, jak mają się do siebie zakresowo obie granice. Jeśli zdolności poznawcze są bardziej ograniczone niż złożoność natury, to teoria ostateczna jest nieosiągalna. Natomiast jeśli są równe lub większe, to możemy mówić potencjalnie o~końcu nauki i~możliwości zaproponowania teorii, która będzie teorią wszystkiego nawet w~sensie homogenicznym. Warto w~tym kontekście zacytować dwa niezwykle optymistyczne zdania Weinberga ze słynnej książki \textit{Sen o~teorii ostatecznej}: ,,Osobiście uważam, że teoria ostateczna istnieje i~że jesteśmy w~stanie ją odkryć. [...] Jednej rzeczy możemy być jednak pewni: odkrycie teorii ostatecznej nie będzie oznaczało kresu nauki'' 
%\label{ref:RNDQ0HYwE4IyO}(Weinberg, 1992, s.~186.190).
\parencite[][s.~186.190]{weinberg_sen_1992}.%

>>>>

\begin{artplenv2auth}{Marek Szydłowski, Paweł Tambor}
	{Czy fizyczne teorie efektywne są wiarygodną strategią osiągnięcia teorii ostatecznej?}
	{Czy fizyczne teorie efektywne są wiarygodną strategią\ldots}
	{Czy fizyczne teorie efektywne są wiarygodną strategią osiągnięcia teorii ostatecznej?}
	{Wydział Filozofii KUL, Uniwersytet Jagielloński Kraków}
	{Are Physical Effective Theories the Reliable Strategy for Achieving Certain Knowledge?}
	{The paper presents the methodology of effective theories as a~strategy used in the process of development of modern physics to reach a~final theory. We present the definition and characteristic features of an effective theory, as well as the answer to the question of whether and what kind of scenario of reaching a~final theory is realized by contemporary physics. We argue that the process of development of physics in the direction of a~final theory is potentially final, i.e. expressible in the conceptual schema of effective theories and as such it is convergent to a~final theory. In each effective theory there are physical constants, however, whose status differs from logical constants. They have a~dimension (length, energy, etc.) and are used to compare physical quantities. The structure of relevant effective theory can be interpreted in the epistemological framework of approximated truth theory. In the case study of cosmological models the sequence of models is convergent to potentially true model. The Standard Cosmological Model is the theory of the structure and dynamics of the Universe.}
	{philosophy and methodology of science, effective theory, final theory, cosmology, approximated truth theories.}
	{%
			{\flushright\subbold{Marek Szydłowski}\\\subsubsectit\small{Uniwersytet Jagielloński}\par}%
			{\flushright\subbold{Paweł Tambor}\\\subsubsectit\small{Katolicki Uniwersytet Lubelski}\par}%
		}

@@

\section{Nomenklatura: teorie efektywne, fundamentalne, ostateczne i~relacje między nimi}
Teoria efektywna posiada przede wszystkim następujące cechy\footnote{Prototypem teorii efektywnej na gruncie fizyki jest efektywna teoria pola
%\label{ref:RNDdsViEs8BX9}(Kim, 1998; Wilson, 1971).
\parencites[][]{kim_wilson_1998}[][]{wilson_renormalization_1971}. %
 W~niniejszej pracy podejmujemy próbę adaptacji pojęcia ,,teoria efektywna'' w~kontekście metodologii i~filozofii nauki.}:
 a)~wyraźnie określony i~ograniczony (najczęściej w~skali energetycznej) zakres stosowania; b) posługuje się parametrami, których natury sama teoria efektywna nie wyjaśnia, co nie przeszkadza w~skutecznym posługiwaniu się teorią; c) posługuje się pojęciami, które mają w~jej ramach status użytecznych fikcji (ze względu na zastosowane idealizacje); d) nie da się do niej stosować predykatu prawdziwości, ze względu na zastosowane idealizacje\footnote{Cechy (c) i~(d) są specyficzne dla analizy Uzana i~nie są powszechnie przyjmowane w~literaturze.}. Jean-Philippe Uzan potwierdza taką intuicję dotyczącą teorii efektywnych; a~zwłaszcza to, że standardowy model kosmologiczny jest taką~właśnie teorią. Model ten z~punktu widzenia teorii fizyki opiera się~na ogólnej teorii względności, elektromagnetyzmie i~fizyce jądrowej. Dość~dobrze da się~wyodrębnić jego ograniczenia w~skali energii. A~zatem, stwierdza Uzan, żeby dobrze rozumieć w~przyjętej skali nasz obserwowalny Wszechświat, nie musimy konstruować kwantowej teorii grawitacji 
%\label{ref:RNDtodLlEAyWN}(Uzan, 2017, s.~109).
\parencite[][s.~109]{uzan_emergent_2017}.%

@@

Obok zaprezentowania kategorii teorii efektywnych i~ostatecznych, celem naszej pracy jest pewna spekulacja nad możliwymi scenariuszami poszukiwania teorii ostatecznej i~propozycja oparcia tych strategii na pojęciu ciągu teorii efektywnych. W~przyjętym rozumieniu teorii ostatecznej, będzie to proces osiągnięcia teorii ostatecznej w~sensie cząstkowym. Cytowany Szynkiewicz w~kontekście własnych analiz wskazuje za Barrowem na cztery scenariusze rozwoju nauki, traktowanej w~aspekcie treściowym
%\label{ref:RNDwI0uwM5FZZ}(Szynkiewicz, 2009, s.~118).
\parencite[][s.~118]{szynkiewicz_teorie_2009}. %
 W~pierwszym (oznaczonym predykatami: natura nieograniczona/zdolności poznawcze nieograniczone) nauka rozwija się w~sposób nieograniczony. W~tym scenariuszu nie ma mowy o~osiągnięciu wiedzy, którą nazwalibyśmy teorią/teoriami ostatecznymi. W~drugim (natura nieograniczona/zdolności poznawcze ograniczone) mamy do czynienia z~dwoma kierunkami rozwoju nauki: a) z~ograniczonymi zdolnościami poznawczymi związany może być (choć nie implikowany przez nie) także nieskończony rozwój nauki, a~zatem nie ma teorii ostatecznej, b) teoria ostateczna jest nieosiągalna, ale z~tego powodu, że ograniczone zdolności poznawcze (także ze względów ekonomicznych lub technicznej niezdolności przetworzenia ogromnej liczby danych) spowodują, że ostatecznie \textit{skazani} będziemy na teorie efektywne. Trzeci scenariusz rozwoju nauki (natura ograniczona/ zdolności poznawcze nieograniczone), jest optymistyczny i~przewiduje możliwość dopracowania się teorii ostatecznej lub przynajmniej zbioru teorii fundamentalnych. Pogląd ten wspierają sukcesy na polu konstruowania teorii kwantowych. Czwarta wizja rozwoju nauki wychodzi z~założenia o~ograniczoności samej natury i~ograniczoności zdolności poznawczych. Możliwe scenariusze zależą od tego, jak mają się do siebie zakresowo obie granice. Jeśli zdolności poznawcze są bardziej ograniczone niż złożoność natury, to teoria ostateczna jest nieosiągalna. Natomiast jeśli są równe lub większe, to możemy mówić potencjalnie o~końcu nauki i~możliwości zaproponowania teorii, która będzie teorią wszystkiego nawet w~sensie homogenicznym. Warto w~tym kontekście zacytować dwa niezwykle optymistyczne zdania Weinberga ze słynnej książki \textit{Sen o~teorii ostatecznej}: ,,Osobiście uważam, że teoria ostateczna istnieje i~że jesteśmy w~stanie ją odkryć. [...] Jednej rzeczy możemy być jednak pewni: odkrycie teorii ostatecznej nie będzie oznaczało kresu nauki'' 
%\label{ref:RNDQ0HYwE4IyO}(Weinberg, 1992, s.~186.190).
\parencite[][s.~186.190]{weinberg_sen_1992}.%
